Parallel Information Set Generation for Kriegspiel

### 2012-07-11 sbac/paper.tex
Several minor changes to the main text of the paper. Still need to work on the section about Information Set Generation.

\begin{abstract}
Information Set Generation is the identification of the set of paths in an imperfect information game tree that are
consistent with a player's observations.  The ability to reason about the possible game history is critical to the
performance of game playing agents.  ISG represent a class of combinatorial search problems which is computationaly intensive but difficult to efficiently parallelize.

In this paper, we address the parallelization of information set generation in the context
of Kriegspiel (partially observable chess).  We implement the algorithm on top of a general purpose combinatorial search
engine and discuss its performance under different load balancing strategies and with different grain-size parameters.
We achieve speedups of over 500 on 1024 processors, far exceeding previous scalability results for game tree search applications.
\end{abstract}

@@

Information Set Generation (ISG) is an operation that game tree search agents
can use to identify nodes in the game tree that are consistent with their
observations~\cite{richards12reasoning}.  Because the operation is relatively
expensive and may be repeatedly applied by a decision-making agent, its
efficiency is a key factor in an agent's overall performance. Information set generation is not the same problem as solving a game tree, although it is arguably a necessary subroutine for quality game play in imperfect information games.

@@

For many specific games, solving the information set generation problem is trivial.  For example, in a poker game, it is easy to
see that the unseen cards held by a player's opponents may be any permutation of the cards not seen by that player
(i.e., hole cards and any revealed community cards).  After the betting rounds, it would not be reasonable to assume
that each of these permutations of unseen cards is equally probable, as betting decisions made by the players up to that
point would be affected by the quality of those players' cards.  But the set of {\em possible} hands for all of the
opponents is easy to conceive and enumerate.

In the game of Kriegspiel, information set generation is much more complicated.  The player knows the opponent's position with
certainty when the game begins, but after the initial move there may be varying levels of uncertainty about the location
of the opponent's pieces.  Unlike poker, it is not possible to simply permute all of the opponent's possible pieces over
all of the squares not occupied by the player's own pieces.  A configuration of pieces for the opponent is only valid if
it can be reached by a legal sequence of moves.  In the general case, finding the nodes in an information set is a
combinatorial search problem.

In this work, we show that ISG for Kriegspiel can be efficiently parallelized. The primary contributions of this work are the following. We describe an approach to efficient parallelization of an application which is not typical of high performance computing domain. We present scalability results for our parallel implementation of information set generation algorithm for the game of Kriegspiel using real game problem instances. Further, we demonstrate the effect of load balancing and grain size on parallel performance of our solution.

The paper is organized as follows.  In Section~\ref{Background}, we review necessary background information, including the rules of kriegspiel and the nature of the information set generation problem.  In Section ~\ref{ParSSSE}, we describe the {\sc Charm++} Parallel State Space Search Engine (ParSSSE) platform.  We then provide our experimental results in Section~\ref{Results}.  Related work is discussed in Section~\ref{Related}.  We conclude with a discussion of future work.

%\Section{Information Set Generation}
\Section{Background: Kriegspiel}
Information Set Generation is a widely adopted operation used in game trees search. Here, we present our techniques and results in the context of a particular game - Kriegspiel. However, the techniques and framework presented in this paper is general and can be applied to other imperfect games for generating information set.
%\Section{Background: Kriegspiel}
Kriegspiel, or partially observable chess,  was invented by Henry Michael Temple in 1899.  Historically, the
game has required three parties: two competing players and one referee.  Today,
players often compete over a computer network, with the role of the referee
being automated.  One player sees and controls only the white pieces; the
other, black.  The referee track of both sets of pieces.  The players alternate
moves as in a regular chess game, but instead of openly declaring their moves,
they secretly communicate move requests to the referee.  The referee, who knows
the location of all pieces, checks the legality of the move.

If the move is legal, the referee makes executes the move and announces that a
legal move has been made and that it is the other player's turn to move.  If
the move is illegal because it is blocked by an opponent's piece or would place
or leave the active player in check, the referee announces, ``No.''  The player
then continues, without penalty, to attempt moves until one finally succeeds.
If a player has no legal moves because of stalemate or checkmate, the referee
announces the game's result.

The referee makes other announcements besides declaring move illegal, and there
are several variations of the game that differ only in the nature of these
announcements.  All declarations by the referee are made to both players.  In
particular, a player will hear and know if her opponent has attempted an
illegal move, and will thus know that her opponent's pieces are configured in
such a way as to allow at least one attemptable move that is blocked or would
leave him in check.

The referee announces when a player is in check. In the case of the capture,
the referee announces the location of the piece that was captured but not its
identity.  Because there is no penalty for making illegal moves, it is often to
a player's advantage to attempt to make moves which are likely to be illegal,
as hearing such declarations from the referee can provide important information
to a player about the location of the opponent's pieces.  In particular, since
a diagonal move by a pawn is legal only the case where such a move would
capture an opponent's piece, it could potentially be profitable to attempt all
or many pawn captures on every turn.  In order to speed the game up, the
referee declares, at the beginning of a player's turn, all locations where a
pawn capture is possible.

@@

%\begin{figure}
%\begin{verbatim}
%1. d4 {(:)}
%   a5 {(:)}
%2. Bg5 {(:)}
%   b6 {(:)}
%3. Nc3 {(:Bd8)}
%   c5 {(:)}
%4. d5 {(:Bd8;Tc5)}
%   Na6 {(:)}
%5. d6 {(:)}
%   f6 {(:Td6)}
%6. e4 {(:)}
%   fxg5 {(Xg5:Td6;Tg5)}
%7. Bc4 {(:)}
%   exd6 {(Xd6:Td6)}
%8. Qd5 {(:)}
%   Qc7 {(:)}
%9. Qf7+ {(:)} 
%   d8 {:g6,f7,e7}
%10. Qxf8++ {(1-0:)}
%\end{verbatim}
%\label{listing}
%\caption{Listing of a kriegspiel game resulting in checkmate by white after 10 moves.}
%\end{figure}

\Section{Information Set Generation}
%\subsection{Information Set Generation}
The concept of information sets is perhaps best explained through an example. 
Figure~\ref{listing} and show a full listing and the associated pictorial progression of an actual
kriegspiel game~\cite{li94chess}.  We have adapted the notation used by Wolfe for Berkeley Kriegspiel.  Odd (resp. even) moves are for white (resp. black).  The actual move is shown
first followed by a list of referee announcements.  Captures announcements are prefixed with an X and give the location
of the captured piece.  Similarly, pawn tries announcements are prefixed with T.  Attempted moves that were ultimately
declared illegal are shown in a list following the `:'.  For example at move 9, (Figure~\ref{fig:9W}--\ref{fig:9B}),
black attempts to block a potential threat along the diagonal by moving a pawn to f6, then attempts to escape or capture
a threat at f7.  When both of these attempts fail, black can infer that his king is being threatened by a protected
bishop or queen at f7.  The attempt to move to e7 is black's way of finding out whether the threatening piece is a queen
or bishop.  When this move fails, black knows it is white's queen and ultimately retreats to d8.  Note that the failed
moves by white on turns 3 and 4 are a result of being blocked by the pawn at e2.

The listing in Figure~\ref{listing} corresponds to the referee's view of the game and would be useful for
post-game analysis or commentary.  During the game, players would not have access to the full information.  The
transcript from white's perspective is shown in Figure~\ref{filteredlisting}.  Note that black's actual moves have been
replaced with ``??'' and lists of black's illegal moves have been replaced with the {\em number} of illegal attempted
moves.  For example, on move 9, black attempted three illegal moves.
\begin{figure}
\begin{minipage}[b]{0.35\linewidth}
\centering
\small
\begin{verbatim}
1. d4 {(:)}
   ?? {(:0)}
2. Bg5 {(:)}
   ?? {(:0)}
3. Nc3 {(:Bd8)}
   ??  {(:0)}
4. d5 {(:Bd8;Tc5)}
   ??  {(:0)}
5. d6 {(:)}
   ?? {(:0,Td6)}
\end{verbatim}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
\small
\centering
\begin{verbatim}
6. e4 {(:)}
   fxg5 {(Xg5:Td6;Tg5)}
7. Bc4 {(:)}
   ?? {(Xd6:Td6)}
8. Qd5 {(:)}
   ?? {(:0)}
9. Qf7+ {(:)} 
   ?? {:0}
10. Qxf8++ {(1-0:)}
\end{verbatim}
\end{minipage}
\vspace{-0.1in}
\caption{Listing of a kriegspiel game resulting in checkmate by white after 10 moves.}
\label{filteredlisting}
\vspace{-0.2in}
\end{figure}

@@

%\begin{table}
%\centering
%\begin{tabular}{rrrrr}
% & \multicolumn{2}{c}{\bf White} & \multicolumn{2}{c}{\bf Black} \\
%{\bf Ply} & {\bf Size} & {\bf Time} & {\bf Size} & {\bf Time} \\
%1 & 18 & 0.000 & 20 & .000 \\
%2 & 18 & 0.000 & 19 & .000 \\
%3 & 18 & 0.000 & 404 & .008\\
%4 & 279 & 0.008 & 401 & .024\\
%5 & 242 & 0.028 & 1472 & .044\\
%6 & 216 & 0.052 & 155 & .096\\
%7 & 176 & 0.064 & 293 & .116\\
%8 & 3406 & 0.092 & 158 & .128\\
%9 & 2191 & 0.288 & 118 & .148\\
%10 & 1423 & 0.508 & 25 & .152\\
%11 & 1363 & 0.596 & 798 & .164\\
%12 & 1416 & 0.744 & 518 & .192\\
%13 & 1416 & 0.836 & 13564 & .304\\
%14 & 3440 & 0.97 & 12394 & .776\\
%15 & 3428 & 1.25 & 343652 & 3.75\\
%16 & 50521 & 1.82 & 320704 & 17.3\\
%17 & 43192 & 5.91 & 490162 & 99.9\\
%18 & 26128 & 7.94 & 3792 & 119\\
%19 & 19061 & 9.92 & 14836 & 121\\
%\end{tabular}
%\caption{Solution counts and running times for a sample kriegspiel game}
%\label{bothtimes}
%\end{table}

%Table~\ref{sampletimes}
%\begin{table}
%\begin{tabular}{ccc}
%1 & 20 & .000\\
%2 & 19 & .000\\
%3 & 404 & .008\\
%4 & 401 & .024\\
%5 & 1472 & .044\\
%6 & 155 & .096\\
%7 & 293 & .116\\
%8 & 158 & .128\\
%9 & 118 & .148\\
%10 & 25 & .152\\
%11 & 798 & .164\\
%12 & 518 & .192\\
%13 & 13564 & .304\\
%14 & 12394 & .776\\
%15 & 343652 & 3.75\\
%16 & 320704 & 17.3\\
%17 & 490162 & 99.9\\
%18 & 3792 & 119\\
%19 & 14836 & 121 \\
%\end{tabular}
%\caption{Solution counts and running times for a sample kriegspiel game}
%\label{blacktimes}
%\end{table}

\textbf{The Information Set Generation Algorithm}
(ISG) can be formulated as a combinatorial search problem.  The input to the algorithm is a list
of observations for one player in a kriegspiel game.

@@

\Section{Parallel Implementation using Charm++ and ParSSSE}\label{ParSSSE}
We built our search engine on top of the {\sc Charm++} run-time
system for parallel processing~\cite{kale93charm,kale09charm}.  {\sc Charm++}
is an extension to the C++ programming language that provides
machine-independent infrastructure for parallel computation.  The language and
its accompanying runtime system have been ported to many shared-memory and
distributed-memory platforms.

In contrast to other popular parallel programming frameworks such as {\sc MPI}
and {\sc OpenMP}, the {\sc Charm++} language is based on object-oriented
programming principles.  The programmer is responsible to decompose its problem
into a collection of objects that represent elements of work to be done.
Typically the number of work objects far exceeds the number of processing
elements (cores) available.  These objects are defined so as to be able to be
migrated between processors {\em during execution}, in order to improve the
overall load balancing.

Work objects, known as {\em chares}, communicate with each other via
asynchronous message-passing.  Objects can invoke {\em entry methods} on other
objects, which are executed according to a schedule set by the {\sc Charm++}
runtime system.  The runtime is responsible to interleave computation and
communication tasks in order to maximize efficiency.  To facilitate this, each
processor maintains an {\em incoming queue} of messages targeted to chares that
it currently holds.  A corresponding {\em outgoing queue} collects messages
generated by objects on the processor that need to be sent to objects assigned
to other processors.

Some work tasks may inherently require more computation than others, but the
amount of computation associated with each piece of work is not always known
before execution.  In tree search, for example, a piece of work may be defined
as the computation of a minimax value on a node in the tree using alpha-beta
pruning.  A natural decomposition is to assign different nodes to different
processors for evaluation.  Because some nodes will be pruned at shallower
levels of the tree than others, the amount of work performed at each subtree can
vary considerably.  In a traditional parallel framework, this would result in
many processors idling while a small number of processors execute the larger
workloads. By allowing the {\sc Charm++} runtime system to schedule a large
number of small pieces of work, better adaptive load balancing is achieved.

%\subsection{The {\sc Charm++} Parallel State Space Search Engine}
Sun \etal have developed the Parallel State Space Search Engine ({\sc ParSSSE})
within the {\sc Charm++} framework, which exposes an API that is useful for a
variety of graph-based and tree-based search problems~\cite{sun11adaptive}. In
{\sc ParSSE} the fundamental unit of work is the processing of a node.
Generally, this means identifying whether the node corresponds to a solution
and/or generating descendant nodes and spawning work tasks related to them. A
task is {\em terminal} if it spawns no children.

A {\em grainsize control} parameter specifies a bound on the size of a piece of
work that can be passed along to the runtime system for scheduling.  Once a
task has been subdivided into an appropriate number of subtasks, those tasks
run to completion on a single processor without preemption and without generating
any additional parallel tasks.  If the grainsize is too large, then the number
of pieces of work to be done may be lower than the number of processors
available, forcing some processors to idle.  If the grainsize is too small,
then the runtime system is flooded with a large number of trivial tasks, and
the system gets bogged down by the overhead of creating and scheduling tasks
instead of actually doing them.  {\sc ParSSE} supports an adaptive grainsize
management, so that the threshold for sequential processing can be changed
during the execution of the program, depending on the properties of the problem
and/or the realtime performance analysis of the runtime system.

{\sc ParSSE} currently supports two different load balancing strategies.  The
first is the randomized strategy.  In this approach, each newly created object
is assigned to a random processor.  Generally, if a large enough number of 
chares is created, the system will achieve an approximately uniform
distribution of work across processors.  The approach is also appealing because
of its simplicity. However, randomization can sometimes lead to inefficiencies
from parallel overhead.  For a program running on $p$ processors, the
probability that a piece of work is assigned to a processor other than the
one that created it is $1-1/p$.  As the number of processors gets large, this
probability approaches 1, which means that the parallel overhead related to the
creation and scheduling of a new task is incurred for almost every piece of new
work.

A second load balancing strategy is known as {\em work stealing}.  In this
approach, all new tasks are inserted in the message queue of the processor that
spawns them.  Whenever a processor is idle it becomes a {\em thief}.  A thief
randomly selects another processor as its {\em victim} and sends a message to
that processor requesting work.  When the victim receives the message, it
removes the work with the highest priority from its own queue and sends it to
the thief.  If it has no work available, it sends a negative acknowledgment
message to the thief, which causes the thief to look for another victim.

The advantage of the work stealing strategy is that the message-passing
overhead is only incurred if at least one processor is idling.  Furthermore,
because victims respond by sending the highest priority work, the most
important tasks are more likely to get done sooner.

@@

\Section{Performance Results}\label{Results}
In order to understand the effect of various parameters such as grain size and load balancing, we first evaluated the scalability of our implementation using a randomly selected problem instance. Next, we show that our implementation performs equally well for instances from actual games.
\subsection{Analysis using Random Instance}
The performance results reported in
this section were obtained by running our program on Blueprint with up to 1024 processors. TODO: Blueprint H/w or citation

Table~\ref{speedups} shows the performance results for our algorithm on a randomly selected problem instance in which
the size of the information set was about 4.5 million.  This instance is not necessarily representative of all games or
states of the game, but it was not cherry-picked in any way. The table shows that the problem is indeed highly
parallizable.  The speedups are nearly linear for up  to 32 processors, and even on 1024 processors, we get a good
speedup.  In this particular instance, the sequential algorithm takes almost 25 minutes to run.  But when parallelized
on 1024 processors---perfectly reasonable resource usage on modern hardware---the search takes less than three seconds.
This is a speedup of 576 times!  It might not be reasonable to use the algorithm at all if it takes 25 or more to run
for some turns.  But if the cost is only a few seconds, it becomes entirely feasible.

\begin{table}
\centering
\begin{tabular}{rrr}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}
\vspace{-0.1in}
\caption{Speedups for a randomly selected problem instance on 1--1024 processors.  The input observations were from the
10th ply of the game tree and the size of the information set in this instance is 4,495,121.}
\label{speedups}
\vspace{-0.3in}
\end{table}

\subsubsection{Effect of Load Balancing}
\begin{figure}[h]
\centering
\includegraphics[width=0.9\columnwidth]{plots/3schemes.pdf}
\vspace{-0.1in}
\caption{Effect of load balancing strategy}
\label{3schemes}
\vspace{-0.2in}
\end{figure}

In our next round of experiments, we set out to compare a few different load-balancing strategies for the problem.  We
used the random, work stealing, and neighborhood balancing strategies that are available in the general purpose search
engine.  These results are shown in Figure~\ref{3schemes}.  Our results show that there was actually very little
variation in performance across the different strategies.  We did notice that across all of our experiments, the work
stealing approach appeared to be consistently---if only marginally---better than the others.  Our hypothesis here is
that perhaps the nodes are rarely starved for work.  In other words, perhaps the very nature of the problem means that
the load is naturally well-balanced and that the load-balancer rarely needs to do a significant amount of
redistribution.  This is probably also related to the fact that we are doing an all-solutions search.

>>>>

\begin{abstract}
Information Set Generation is the identification of the set of paths in an
imperfect information game tree that are consistent with a player's
observations.  The ability to reason about the possible game history is
critical to the performance of game-playing agents.  ISG represent a class of
combinatorial search problems which is computationaly intensive but difficult
to efficiently parallelize.

In this paper, we address the parallelization of information set generation in
the context of Kriegspiel (partially observable chess).  We implement the
algorithm on top of a general purpose combinatorial search engine and discuss
its performance under different load balancing strategies and with different
grainsize parameters.  We achieve speedups of over 500 on 1,024 processors, far
exceeding previous scalability results for game tree search applications.
\end{abstract}

@@

Information Set Generation (ISG) is an operation that game tree search agents
can use to identify nodes in the game tree that are consistent with their
observations~\cite{richards12information}.  Because the operation is
computationally intensive and may be repeatedly applied by a decision-making
agent, its efficiency is a key factor in an agent's overall performance.

@@

For many specific games, solving the information set generation problem is
trivial.  For example, in a Poker game, the unseen cards held by a player's
opponents may be any permutation of the cards not seen by that player (i.e.,
hole cards and any revealed community cards).  After the betting rounds, it
would not be reasonable to assume that each of these permutations of unseen
cards is equally probable, as betting decisions made by the players up to that
point would be affected by the quality of those players' cards.  But the set of
{\em possible} hands for all of the opponents is easy to conceive and
enumerate.

In the game of Kriegspiel, information set generation is much more complicated.
After the first move by the opponent, a player faces uncertainty about the
location of the opponent's pieces.  Unlike Poker, it is not possible to simply
permute all of the opponent's possible pieces over all of the squares not
occupied by the player's own pieces.  A configuration of pieces for the
opponent is only valid if it can be reached by a legal sequence of moves.  In
the general case, finding the nodes in an information set is a combinatorial
search problem.

In this work, we show that ISG for Kriegspiel can be efficiently parallelized.
The primary contributions of this work are the following. We describe an
approach to efficient parallelization of an application which is not typical of
high performance computing domain. We present scalability results for our
parallel implementation of an information set generation algorithm for the game
of Kriegspiel using problem instances from real games. Further, we demonstrate
the effects of load balancing strategy and grainsize on parallel performance.

The paper is organized as follows.  In Section~\ref{Background}, we review
necessary background material, including the rules of Kriegspiel and the nature
of the information set generation problem.  In Section ~\ref{ParSSSE}, we
describe the {\sc Charm++} Parallel State Space Search Engine (ParSSSE)
platform.  We then provide our experimental results in Section~\ref{Results}.
Related work is discussed in Section~\ref{Related}.  We conclude with a
discussion of future work.

%\Section{Information Set Generation}
\Section{Kriegspiel}
Information Set Generation is the process of identifying the nodes in an
imperfect information game tree that are consistent with a player's
observations~\cite{richards12information}. Here, we present our techniques and
results in the context of a particular game: Kriegspiel~\cite{li94chess}.
However, the techniques and framework presented in this paper are general and
can be applied to ISG in other imperfect information games.

%\Section{Background: Kriegspiel}
Kriegspiel, or partially observable chess,  was invented by Henry Michael
Temple in 1899.  Historically, the game has required three parties: two
competing players and one referee.  Today, players often compete over a
computer network, with the role of the referee being automated.  One player
sees and controls only the white pieces; the other, black.  The referee tracks
both sets of pieces.  Instead of openly declaring their moves, players secretly
communicate move requests to the referee.  The referee, who knows the location
of all pieces, assesses the legality of the move according to the standard
rules of chess.

If the move is legal, the referee executes the move and announces that a
legal move has been made and that it is the other player's turn to move.  If
the move is illegal because it is blocked by an opponent's piece or would place
or leave the active player in check, the referee announces, ``No.''  The player
then continues, without penalty, to attempt moves until one finally succeeds.
If a player has no legal moves because of stalemate or checkmate, the referee
announces the game's result.

All declarations by the referee are made to both players.  In
particular, a player will hear and know if her opponent has attempted an
illegal move, and will thus know that her opponent's pieces are configured in
such a way as to allow at least one attemptable move that is blocked or would
leave him in check.

The referee makes other announcements besides declaring move illegal, and there
are several variations of the game that differ only in the nature of these
announcements.  In our implementation, the referee announces when a player is
in check, when a piece has been captured, and whenever a pawn may capture an
opponent's piece.  In the case of a capture, the referee announces the square
in which the capture occurs but not the identity of the captured piece.

@@

%\begin{figure}
%\begin{verbatim}
%1. d4 {(:)}
%   a5 {(:)}
%2. Bg5 {(:)}
%   b6 {(:)}
%3. Nc3 {(:Bd8)}
%   c5 {(:)}
%4. d5 {(:Bd8;Tc5)}
%   Na6 {(:)}
%5. d6 {(:)}
%   f6 {(:Td6)}
%6. e4 {(:)}
%   fxg5 {(Xg5:Td6;Tg5)}
%7. Bc4 {(:)}
%   exd6 {(Xd6:Td6)}
%8. Qd5 {(:)}
%   Qc7 {(:)}
%9. Qf7+ {(:)} 
%   d8 {:g6,f7,e7}
%10. Qxf8++ {(1-0:)}
%\end{verbatim}
%\label{listing}
%\caption{Listing of a Kriegspiel game resulting in checkmate by white after 10 moves.}
%\end{figure}

\Section{Information Set Generation}
%\subsection{Information Set Generation}
The concept of information sets is perhaps best explained through an example. 
Figure~\ref{listing} and show a full listing and the associated pictorial progression of an actual
Kriegspiel game~\cite{li94chess}.  We have adapted the notation used by Wolfe for Berkeley Kriegspiel.  Odd (resp. even) moves are for white (resp. black).  The actual move is shown
first followed by a list of referee announcements.  Captures announcements are prefixed with an X and give the location
of the captured piece.  Similarly, pawn tries announcements are prefixed with T.  Attempted moves that were ultimately
declared illegal are shown in a list following the `:'.  For example at move 9, (Figure~\ref{fig:9W}--\ref{fig:9B}),
black attempts to block a potential threat along the diagonal by moving a pawn to f6, then attempts to escape or capture
a threat at f7.  When both of these attempts fail, black can infer that his king is being threatened by a protected
bishop or queen at f7.  The attempt to move to e7 is black's way of finding out whether the threatening piece is a queen
or bishop.  When this move fails, black knows it is white's queen and ultimately retreats to d8.  Note that the failed
moves by white on turns 3 and 4 are a result of being blocked by the pawn at e2.

The listing in Figure~\ref{listing} corresponds to the referee's view of the game and would be useful for
post-game analysis or commentary.  During the game, players would not have access to the full information.  The
transcript from white's perspective is shown in Figure~\ref{filteredlisting}.  Note that black's actual moves have been
replaced with ``??'' and lists of black's illegal moves have been replaced with the {\em number} of illegal attempted
moves.  For example, on move 9, black attempted three illegal moves.
\begin{figure}
\begin{minipage}[b]{0.35\linewidth}
\centering
\small
\begin{verbatim}
1. d4 {(:)}
   ?? {(:0)}
2. Bg5 {(:)}
   ?? {(:0)}
3. Nc3 {(:Bd8)}
   ??  {(:0)}
4. d5 {(:Bd8;Tc5)}
   ??  {(:0)}
5. d6 {(:)}
   ?? {(:0,Td6)}
\end{verbatim}
\end{minipage}
\begin{minipage}[b]{0.25\linewidth}
\small
\centering
\begin{verbatim}
6. e4 {(:)}
   fxg5 {(Xg5:Td6;Tg5)}
7. Bc4 {(:)}
   ?? {(Xd6:Td6)}
8. Qd5 {(:)}
   ?? {(:0)}
9. Qf7+ {(:)} 
   ?? {:0}
10. Qxf8++ {(1-0:)}
\end{verbatim}
\end{minipage}
\vspace{-0.1in}
\caption{Listing of a Kriegspiel game resulting in checkmate by white after 10 moves.}
\label{filteredlisting}
\vspace{-0.2in}
\end{figure}

@@

%\begin{table}
%\centering
%\begin{tabular}{rrrrr}
% & \multicolumn{2}{c}{\bf White} & \multicolumn{2}{c}{\bf Black} \\
%{\bf Ply} & {\bf Size} & {\bf Time} & {\bf Size} & {\bf Time} \\
%1 & 18 & 0.000 & 20 & .000 \\
%2 & 18 & 0.000 & 19 & .000 \\
%3 & 18 & 0.000 & 404 & .008\\
%4 & 279 & 0.008 & 401 & .024\\
%5 & 242 & 0.028 & 1472 & .044\\
%6 & 216 & 0.052 & 155 & .096\\
%7 & 176 & 0.064 & 293 & .116\\
%8 & 3406 & 0.092 & 158 & .128\\
%9 & 2191 & 0.288 & 118 & .148\\
%10 & 1423 & 0.508 & 25 & .152\\
%11 & 1363 & 0.596 & 798 & .164\\
%12 & 1416 & 0.744 & 518 & .192\\
%13 & 1416 & 0.836 & 13564 & .304\\
%14 & 3440 & 0.97 & 12394 & .776\\
%15 & 3428 & 1.25 & 343652 & 3.75\\
%16 & 50521 & 1.82 & 320704 & 17.3\\
%17 & 43192 & 5.91 & 490162 & 99.9\\
%18 & 26128 & 7.94 & 3792 & 119\\
%19 & 19061 & 9.92 & 14836 & 121\\
%\end{tabular}
%\caption{Solution counts and running times for a sample Kriegspiel game}
%\label{bothtimes}
%\end{table}

%Table~\ref{sampletimes}
%\begin{table}
%\begin{tabular}{ccc}
%1 & 20 & .000\\
%2 & 19 & .000\\
%3 & 404 & .008\\
%4 & 401 & .024\\
%5 & 1472 & .044\\
%6 & 155 & .096\\
%7 & 293 & .116\\
%8 & 158 & .128\\
%9 & 118 & .148\\
%10 & 25 & .152\\
%11 & 798 & .164\\
%12 & 518 & .192\\
%13 & 13564 & .304\\
%14 & 12394 & .776\\
%15 & 343652 & 3.75\\
%16 & 320704 & 17.3\\
%17 & 490162 & 99.9\\
%18 & 3792 & 119\\
%19 & 14836 & 121 \\
%\end{tabular}
%\caption{Solution counts and running times for a sample Kriegspiel game}
%\label{blacktimes}
%\end{table}

\textbf{The Information Set Generation Algorithm}
(ISG) can be formulated as a combinatorial search problem.  The input to the algorithm is a list
of observations for one player in a Kriegspiel game.

@@

\Section{Parallel Implementation using Charm++ and ParSSSE}\label{ParSSSE}
We implemented our search engine on top of the {\sc Charm++} run-time
system for parallel processing~\cite{kale93charm,kale09charm}.  {\sc Charm++}
is an extension to the C++ programming language that provides
machine-independent infrastructure for parallel computation.  The language and
its accompanying runtime system have been ported to many shared-memory and
distributed-memory platforms.

In contrast to other popular parallel programming frameworks such as {\sc MPI}
and {\sc OpenMP}, the {\sc Charm++} language is based on object-oriented
programming principles.  The programmer is responsible to decompose a problem
into a collection of objects that represent elements of work to be done.
Typically the number of these work objects, called {\em chares}, far exceeds
the number of processing elements (cores) available.  Chares are defined
so as to be able to be migrated between processors {\em during execution}, in
order to improve the overall load balancing.

Chares communicate with each other via
asynchronous message passing.  They can invoke {\em entry methods} on other
objects, which are executed according to a schedule set by the {\sc Charm++}
runtime system.  The runtime is responsible to interleave computation and
communication tasks in order to maximize efficiency.  To facilitate this, each
processor maintains an {\em incoming queue} of messages targeted to chares that
it currently holds.  A corresponding {\em outgoing queue} collects messages
generated by objects on the processor that need to be sent to chares assigned
to other processors.

Some chares may inherently require more computation than others, but the amount
of computation associated with each piece of work is not always known before
execution.  In tree search, for example, a piece of work may be defined as the
computation of a minimax value on a node in the tree using alpha-beta
pruning~\cite{knuth75analysis,baudet78analysis}.  A natural decomposition is to
assign different nodes to different processors for evaluation.  Because some
nodes will be pruned at shallower levels of the tree than others, the amount of
work performed at each subtree can vary considerably.  In a traditional
parallel framework, this would result in many processors idling while a small
number of processors execute the larger workloads. By allowing the {\sc
Charm++} runtime system to schedule a large number of small pieces of work,
better adaptive load balancing is achieved.

%\subsection{The {\sc Charm++} Parallel State Space Search Engine}
Sun \etal have developed the Parallel State Space Search Engine ({\sc ParSSSE})
within the {\sc Charm++} framework, which exposes an API that is useful for a
variety of graph-based and tree-based search problems~\cite{sun11adaptive}. In
{\sc ParSSSE} the fundamental unit of work is the processing of a node.
Generally, this means identifying whether the node corresponds to a solution
and/or generating descendant nodes and spawning work tasks related to them. A
task is {\em terminal} if it spawns no children.

A {\em grainsize control} parameter specifies a bound on the size of a piece of
work that can be passed along to the runtime system for scheduling.  Once a
task has been subdivided into an appropriate number of subtasks, those tasks
run to completion on a single processor without preemption and without generating
any additional parallel tasks.  If the grainsize is too large, then the number
of pieces of work to be done may be lower than the number of processors
available, forcing some processors to idle.  If the grainsize is too small,
then the runtime system is flooded with a large number of trivial tasks, and
the system gets bogged down by the overhead of creating and scheduling tasks
instead of actually doing them.  {\sc ParSSSE} supports adaptive grainsize
management, so that the threshold for sequential processing can be changed
during the execution of the program, depending on the properties of the problem
and/or the realtime performance analysis of the runtime system.

{\sc ParSSSE} currently supports two load balancing strategies.  The
first is the randomized strategy.  In this approach, each newly created object
is assigned to a random processor.  Generally, if a large enough number of 
chares is created, the system will achieve an approximately uniform
distribution of work across processors.  The approach is also appealing because
of its simplicity. However, randomization can sometimes lead to inefficiencies
from parallel overhead.  For a program running on $p$ processors, the
probability that a chare is assigned to a processor other than the
one that created it is $1-1/p$.  As the number of processors gets large, this
probability approaches 1, which means that the parallel overhead related to the
creation and scheduling of a new task is incurred for almost every piece of new
work.

The second load balancing strategy is {\em work stealing}.  In this
approach, all new tasks are inserted in the message queue of the processor that
spawns them.  Whenever a processor is idle it becomes a {\em thief}.  A thief
randomly selects another processor as its {\em victim} and sends a message to
that processor requesting work.  When the victim receives the message, it
removes the work with the highest priority from its own queue and sends it to
the thief.  If it has no work available, it sends a negative acknowledgment
message to the thief, which causes the thief to look for another victim.

The advantage of the work stealing strategy is that the message-passing
overhead is only incurred if at least one processor is idling.  Furthermore,
because victims respond by sending the highest priority work, the most
important tasks are scheduled sooner.

@@

\Section{Performance Results}\label{Results}
In order to understand the effect of various parameters such as grainsize and
load balancing strategy, we first evaluated the scalability of our
implementation using a randomly selected problem instance. Next, we show that
our implementation performs equally well for instances from actual games.

\subsection{Analysis using Random Instance} 
The performance results reported in this section were obtained by running our
program on Blueprint with up to 1,024 processors. TODO: Blueprint H/w or
citation

Table~\ref{speedups} shows the performance results for our algorithm on a
randomly selected problem instance in which the size of the information set was
about 4.5 million.  This instance is not necessarily representative of all
games or states of the game, but it was not cherry-picked in any way. The table
shows that the problem is indeed highly parallizable. The speedups are nearly
linear for up to 32 processors, and the speedups continue to increase even up
to 1,024 processors. In this particular instance, the sequential algorithm
takes almost 25 minutes to run.  But when parallelized on 1,024
processors---perfectly reasonable resource usage on modern hardware---the
search takes less than three seconds.  This is a speedup of 576 times! In a
tournament where each player is allotted an average of five minutes per move,
an ISG operation that requires 25 minutes or more would be prohibitively
expensive.  On the other hand, if the operation can be done in under three
seconds, ISG can be used as a subroutine in the decision-making process
(see~\cite{parker05game}).

\begin{table}
\centering
\begin{tabular}{rrr}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}
\vspace{-0.1in}

\caption{Speedups for a randomly selected problem instance on 1--1,024
processors.  The input observations were from the 10th ply of the game tree and
the size of the information set in this instance is 4,495,121.}

\label{speedups}
\vspace{-0.3in}
\end{table}

\subsubsection{Effect of Load Balancing}
\begin{figure}[h]
\centering
\includegraphics[width=0.9\columnwidth]{plots/3schemes.pdf}
\vspace{-0.1in}
\caption{Effect of load-balancing strategy}
\label{3schemes}
\vspace{-0.2in}
\end{figure}

In our next round of experiments, we set out to compare load-balancing
strategies for the problem.  We used the random, work stealing, and
neighborhood balancing strategies that are available in the general purpose
search engine.  These results are shown in Figure~\ref{3schemes}.  Our results
show that there was actually little variation in performance across the
different strategies.  Across all of our experiments, the work stealing
approach appeared to be consistently---if only marginally---better than the
others.  Our hypothesis here is that the nodes are rarely starved for work.  In
other words, perhaps the very nature of the problem means that the load is
naturally well-balanced and that the load-balancer rarely needs to do a
significant amount of redistribution.  This is probably also related to the
fact that we are doing an all-solutions search.
